Sequence Specific Modeling of \emph{E.~coli} Cell-Free Protein Synthesis

### 2017-05-19 Paper_v1.tex
oxphos_yield and edits

Cell-free simulations of the deGFP titer for a range of plasmid concentrations were consistent with experimental measurements (Fig.~\ref{fig:network}D).
The titer at each plasmid concentration was calculated by multiplying the deGFP synthesis flux by the active time of production, approximately 8 hours in TXTL 2.0 \cite{Garamella:2016aa}.
The mean of the ensemble (calculated by sampling the uncertainty in the model parameters) captured the saturation of deGFP production as a function of plasmid concentration (R\textsuperscript{2} = 0.97).
However, while the mean and 95\% confidence estimate of the ensemble were consistent with measured deGFP levels, the model underpredicted the deGFP titer at the saturating plasmid concentration of 5 nM.
These results, in combination with the time-dependent CAT and deGFP simulations, validated our modeling approach, which required very few adjustable parameters.
It also showed that the sequence specific template reactions and metabolic network were sufficient to predict the production of different proteins under different promoters and cell-free systems.
Next, we used our modeling approach to better understand the theoretical performance limits of CFPS.

@@

\subsection{Global sensitivity analysis}
To better understand the effect of substrate utilization and the transcription/translation parameters on CFPS performance, we performed global sensitivity analysis on the productivity, energy efficiency and carbon yield for the CAT protein (Fig.~\ref{fig:SI} and Fig. 1 in Supporting Information).
Surprisingly, RNAP and ribosome abundance had only a modest effect; the translation elongation rate had the largest effect on protein productivity.
On the other hand, oxygen and substrate consumption had the largest and second-largest influences on the energy efficiency and carbon yield, respectively.
The significance of transcription/translation parameters was robust to amino acid supplementation, with the translation rate being the most sensitive across all cases (Fig.~\ref{fig:SI}A).
This suggested that the translation elongation rate, and not transcriptional parameters, controlled productivity.
Underwood and coworkers showed that an increase in ribosome levels did not significantly increase protein yields or rates; however, adding elongation factors increased protein synthesis rates by 27\% \cite{2005_underwood_biotech}.
In addition, Li et al. increased the productivity by 5-fold of firefly luciferase in PURE CFPS by first improving the rate-limiting step, translation, followed by transcription by adjusting elongation factors, ribosome recycling factor, release factors, chaperones, BSA, and tRNAs \cite{2014_li_PlosOne}.
In examining substrate utilization, glucose consumption was not important for productivity in the presence of amino acid supplementation.
However, its importance increased significantly when amino acids were not available.
On the other hand, amino acid consumption was only sensitive when amino acids synthesis reactions were blocked, as it was the only source of amino acids for CAT synthesis.
The oxygen consumption rate was the most important factor controlling the energy efficiency of cell-free protein synthesis (Fig.~\ref{fig:SI}B).
In the model, we assumed that ATP could be produced by both substrate level and oxidative phosphorylation.
Jewett and coworkers reported that oxidative phosphorylation still operated in cell-free systems, and that the protein titer decreased from 1.5-fold to 4-fold when oxidative phosphorylation reactions were inhibited in pyruvate-powered CFPS \cite{Jewett:2008aa}.
However, it is unknown how active oxidative phosphorylation is in a glucose-powered cell-free system.
Moreover, the connection between oxidative phosphorylation activity and other performance metrics, such as carbon yield, is also unclear.

@@

To investigate the connection between carbon yield and oxidative phosphorylation further, we calculated the optimal CAT carbon yield as a function of the oxidative phosphorylation flux (Fig.~\ref{fig:oxphos_yield}).
Oxidative phosphorylation had a strong effect on the carbon yield, both with and without amino acid supplementation.
In the presence of amino acid supplementation, the carbon yield ranged from 20\% to approximately 60\%, depending on the oxidative phosphorylation flux.
However, without amino acid supplementation, the carbon yield dropped to approximately 10\%, and reached a maximum of 50\%.
In the absence of supplementation, a lower carbon yield was expected for the same oxidative phosphorylation flux, as glucose was utilized for both energy generation and amino acid biosynthesis.
In all cases, whenever the carbon yield was below its theoretical maximum, there was an accumulation of both acetate and lactate.
The experimental dataset exhibited a mixture of acetate and lactate accumulation during CAT synthesis, which suggested the CFPS reaction was not operating with optimal oxidative phosphorylation activity.
Oxidative phosphorylation is a membrane associated process, while CFPS has no cell membrane.
Jewett et al. hypothesized that membrane vesicles present in the CFPS reaction carry out oxidative phosphorylation \cite{Jewett:2008aa}.
In addition, they were able to enhance the production titer of CAT by 33\% when the reaction was augmented with 10mM phosphate.
However, it is unclear why the additional phosphate increased CAT yield.
Jewett et al. believe it either enhanced oxidative phosphorylation activity or inhibited phosphatase reactions.
The number, size, protein loading, and lifetime of these vesicles remains an open area of study.
Thus, we expect CFPS will have suboptimal oxidative phosphorylation capacity, and it remains an important area for optimization of CFPS performance.

@@

\begin{figure}[t!]
\includegraphics[width=0.7\textwidth]{./figs/Fig-4-OxyPhox-versus-Carbon-Yield.pdf}
\caption{Optimal CAT carbon yield versus oxidative phosphorylation flux calculated across an ensemble (N = 1000) flux balance solutions (points).
Carbon yield versus oxidative phosphorylation flux for amino acid supplementation and \textit{de~novo} synthesis (black),
amino acid supplementation without \textit{de~novo} synthesis (dark grey),
and \textit{de~novo} amino acid synthesis without supplementation (light gray).}
\label{fig:oxphos_yield}
\end{figure}

>>>>

Cell-free simulations of the deGFP titer for a range of plasmid concentrations were consistent with experimental measurements (Fig.~\ref{fig:network}D).
The titer at each plasmid concentration was calculated by multiplying the deGFP synthesis flux by the active time of production, approximately 8 hours in TXTL 2.0 \cite{Garamella:2016aa}.
The mean of the ensemble (calculated by sampling the uncertainty in the model parameters) captured the saturation of deGFP production as a function of plasmid concentration (R\textsuperscript{2} = 0.97).
However, while the mean and 95\% confidence estimate of the ensemble were consistent with measured deGFP levels, the model under predicted the deGFP titer at the saturating plasmid concentration of 5 nM.
These results, in combination with the time-dependent CAT and deGFP simulations, validated our modeling approach, which required very few adjustable parameters.
It also showed that the sequence specific template reactions and metabolic network were sufficient to predict the production of different proteins under different promoters and cell-free systems.
Next, we used our modeling approach to better understand the theoretical performance limits of CFPS.

@@

\subsection{Global sensitivity analysis}
To better understand the effect of substrate utilization and the transcription/translation parameters on CFPS performance, we performed global sensitivity analysis on the productivity, energy efficiency and carbon yield for the CAT protein (Fig.~\ref{fig:SI} and Fig. 1 in Supporting Information).
Surprisingly, RNAP and ribosome abundance had only a modest effect; the translation elongation rate had the largest effect on protein productivity.
On the other hand, oxygen and substrate consumption had the largest and second-largest influences on the energy efficiency and carbon yield, respectively.
The significance of transcription/translation parameters was robust to amino acid supplementation, with the translation rate being the most sensitive across all cases (Fig.~\ref{fig:SI}A).
This suggested that the translation elongation rate, and not transcription parameters, controlled productivity.
Underwood and coworkers showed that an increase in ribosome levels did not significantly increase protein yields or rates; however, adding elongation factors increased protein synthesis rates by 27\% \cite{2005_underwood_biotech}.
In addition, Li et al. increased the productivity by 5-fold of firefly luciferase in PURE CFPS by first improving the rate-limiting step, translation, followed by transcription by adjusting elongation factors, ribosome recycling factor, release factors, chaperones, BSA, and tRNAs \cite{2014_li_PlosOne}.
In examining substrate utilization, glucose consumption was not important for productivity in the presence of amino acid supplementation.
However, its importance increased significantly when amino acids were not available.
On the other hand, amino acid consumption was only sensitive when amino acids synthesis reactions were blocked, as it was the only source of amino acids for CAT synthesis.
The oxygen consumption rate was the most important factor controlling the energy efficiency of cell-free protein synthesis (Fig.~\ref{fig:SI}B).
In the model, we assumed that ATP could be produced by both substrate level and oxidative phosphorylation.
Jewett and coworkers reported that oxidative phosphorylation still operated in cell-free systems, and that the protein titer decreased from 1.5-fold to 4-fold when oxidative phosphorylation reactions were inhibited in pyruvate-powered CFPS \cite{Jewett:2008aa}.
However, it is unknown how active oxidative phosphorylation is in a glucose-powered cell-free system.
Moreover, the connection between oxidative phosphorylation activity and other performance metrics, such as carbon yield, is also unclear.

@@

To investigate the connection between carbon yield and oxidative phosphorylation further, we calculated the optimal CAT carbon yield as a function of the oxidative phosphorylation flux (Fig.~\ref{fig:oxphos_yield}).
We calculated yield across an ensemble of 1000 flux balance solutions by varying the oxygen uptake rate with transcription and translation parameters.
Oxidative phosphorylation had a strong effect on the carbon yield, both with and without amino acid supplementation.
In the presence of amino acid supplementation, the carbon yield ranged from 20\% to approximately 60\%, depending on the oxidative phosphorylation flux.
However, without amino acid supplementation, the carbon yield dropped to approximately 10\%, and reached a maximum of 50\%.
In the absence of supplementation, a lower carbon yield was expected for the same oxidative phosphorylation flux, as glucose was utilized for both energy generation and amino acid biosynthesis.
In all cases, whenever the carbon yield was below its theoretical maximum, there was an accumulation of both acetate and lactate.
The experimental dataset exhibited a mixture of acetate and lactate accumulation during CAT synthesis, which suggested the CFPS reaction was not operating with optimal oxidative phosphorylation activity.
Oxidative phosphorylation is a membrane associated process, while CFPS has no cell membrane.
Jewett et al. hypothesized that membrane vesicles present in the CFPS reaction carry out oxidative phosphorylation \cite{Jewett:2008aa}.
In addition, they were able to enhance the production titer of CAT by 33\% when the reaction was augmented with 10mM phosphate.
However, it is unclear why the additional phosphate increased CAT yield.
Jewett et al. believe it either enhanced oxidative phosphorylation activity or inhibited phosphatase reactions.
The number, size, protein loading, and lifetime of these vesicles remains an open area of study.
Thus, we expect CFPS will have suboptimal oxidative phosphorylation capacity, and it remains an important area for optimization of CFPS performance.

@@

\begin{figure}[t!]
\includegraphics[width=0.7\textwidth]{./figs/Fig-4-OxyPhox-versus-Carbon-Yield.pdf}
\caption{Optimal CAT carbon yield versus oxidative phosphorylation flux calculated across an ensemble (N = 1000) flux balance solutions (points) by varying the oxygen uptake rate with transcription and translation parameters.
Carbon yield versus oxidative phosphorylation flux for amino acid supplementation and \textit{de~novo} synthesis (black),
amino acid supplementation without \textit{de~novo} synthesis (dark grey),
and \textit{de~novo} amino acid synthesis without supplementation (light gray).}
\label{fig:oxphos_yield}
\end{figure}
